Sentence Generation as Planning with Probabilistic LTAG\\

### 2010-05-21 trunk/Papers/pcrisp-10/experiments.tex
experiment section

We estimate our probability models on a training portion of the dataset (96\%) using maximum likelihood estimation.
For the testing set we selected from the remaining 4\% all sentences that were shorter than 16 words (416 sentences). For longer sentences the search time often exceeded 5 minutes.

%{\bf Creating realization problems:} 
For this testing set we compute semantic representation for each sentence, by traversing its derivation tree top-down and collects a conjunction of terms, by instantiating the automatically created semantic representations. Our algorithm introduces a new individual for each substitution node marked with role $r$ and equates it to the `self' role of any tree that adjoins or substitutes into a node marked with $r$.

\subsection{Evaluation Measures}
\label{ssec:eval-measures}
To evaluate the output quality of our statistical generation system we compare the system output $O$ against the reference sentence $R$in the treebank, from which the input representation was generated. We adopt the {\bf generation tree accuracy} measure discussed by \cite{bangalorerambowwhittaker2001}.
This measure is computed by first creating a list of all 'treelets' from the reference derivation tree $D$. A 'treelet' is a subtree, consisting only of a node and its direct children. For each treelet we calculate edit distance, sum the distances over all treelets and then divide by the total length of the reference string.
$$1 - \frac{\sum\limits_{d \in treelets(D)} editDist(O|_d, R|_d)}{\textit{length}(R)},$$
where $D$ is the reference derivation tree and $S|_d$ are the tokens associated with the nodes of treelet $d$ in the order they appear in $S$ (if at all). 
Edit distance is modified to treat insertion-deletion pairs as single movement errors. 
This metric account for the observation that errors that change word order on a local level (i.e. within a treelet) are less severe than errors that swap words between different treelets. Intuitively errors of the latter type violate the continuity of sub-phrases. Because of this property we prefer generation tree accuracy over {\sc blue}, which is overly sensitive to movement errors on an individual sentence level.

@@

\subsection{Planning Systems}
The planning system we used was a reimplementation of {\sc FF} \cite{hoffmannnebel2001}, a heuristic search planner. We extended this planner to take action costs into account when it calculates heuristic values for successor states. We also experimented with the greedy search heuristic used by {\sc spud} \cite{stonedoran1997}.

\subsection{Results}
Table \ref{results} presents our results.

The first two columns confirm our hypothesis that {\sc crisp} does not scale up to the automatically induced treebank. The greedy {\sc spud} heuristic does not generate a single correct sentence, as the combinatorics of the planning domain becomes too large with the treebank induced grammar. 
With the standard {\sc ptag} definition our system manages to generate only very few sentences due to data sparseness. However these sentences are almost identical to the ones found in the test set. Both generalized models ease the data sparseness problem, but output quality drops rapidly. For the unlexicalized model the search problem becomes very hard due to the large planning domain, so that 22\% of the generation problems require more than 5 minutes runtime. As expected the smoothed model offers a compromise between efficiency, output quality and coverage 
\footnote{In our experiments we set the weight between level 1 and 2 to 0.9 and did not consider level 3, after empirically testing different settings. We set a probability threshold to include operators into the planning domain to $10^{-4}$}.

\begin{table}
    \caption{\label{results} Evaluation results.}
    \begin{center}
    \begin{tabular}{|l|l|r|r|r|}
    \multicolumn{5}{c}{\sc crisp}\\ 
    \hline
     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     greedy         & ??? & 0\% & 100\% & 0\%  \\ \hline 
     FF         &  0.66 & 88\% & 0\% & 12\% \\ \hline 
    \multicolumn{5}{c}{\bf planning with {\sc ptag}}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs &  0.99 &  97\% & 0\% & 3\% \\ \hline
     FF         &  0.99 & 71\% & 0\% &  29\% \\ \hline
    \multicolumn{5}{c}{\bf planning with {\sc ptag} (unlexicalized)}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs & 0.74 & 62\% & 16\% & 22\%\\ \hline
     FF & 0.76 & 62\% & 16\% &  22\%\\ \hline
    \multicolumn{5}{c}{\bf planning with {\sc ptag} (smoothed)}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs & 0.78 & 19\% & 74\% & 7\% \\ \hline
     FF & 0.77 & 19\% & 74\% & 7\% \\ \hline
    \end{tabular}
    \end{center}
\end{table}

>>>>

We estimate our probability models on section 1-23 of the converted WSJ using maximum likelihood estimation. We use section 0 as a testing set. However, since the number of {\sc pcrisp} operators grows quadratically, generating long sentences requires too much time to run batch experiments.
We therefore restricted our evaluation to the 416 sentences shorter than 16 words.

%{\bf Creating realization problems:} 
For this testing set we automatically create semantic representation for each sentence, by instantiating the semantic content of each elementary used in its derivation. We use these representations as input for our system and compare the system output against the original sentence.

\subsection{Evaluation Measures}
\label{ssec:eval-measures}
To evaluate the output quality of our statistical generation system we compare the system output $O$ against the reference sentence $R$ in the treebank, from which the input representation was generated. We adopt the {\bf generation tree accuracy} measure discussed by \cite{bangalorerambowwhittaker2001}.
This measure is computed by first creating a list of all 'treelets' from the reference derivation tree $D$. A 'treelet' is a subtree, consisting only of a node and its direct children. For each treelet we calculate edit distance, sum the distances over all treelets and then divide by the total length of the reference string:
$$1 - \frac{\sum\limits_{d \in treelets(D)} editDist(O|_d, R|_d)}{\textit{length}(R)},$$
where $D$ is the reference derivation tree and $S|_d$ are the tokens associated with the nodes of treelet $d$ in the order they appear in $S$ (if at all). 
Edit distance is modified to treat insertion-deletion pairs as single movement errors. 
This metric account for the observation that errors that change word order on a local level (i.e. within a treelet) are less severe than errors that swap words between different treelets. Intuitively errors of the latter type violate the continuity of sub-phrases. Because of this property we prefer generation tree accuracy over {\sc bleu}, which is overly sensitive to movement errors on an individual sentence level.

@@

\subsection{Planning Systems}
For our experiments we use a reimplementation of {\sc FF} \cite{hoffmannnebel2001}, a heuristic search planner. We extend this planner to take action costs into account when it calculates heuristic values for successor states. We also experiment with the greedy heuristic search strategy used by {\sc spud}, which we adapted to problem specific search strategy for the planner.

\subsection{Results}
Table \ref{results} presents our results. In each experiment we count how many sentences in the test set produce an output at all (column `generated') and for how many the planner requires more than 5 minutes (column `timeout'. The search is terminated in this case). Some sentences cannot be realized at all (column `fail'), due to data sparseness problems. In such cases either no elementary tree with the required lexical content was seen in the training data, or the model assigns a zero probability to a required adjunction or substitution operation. We compute the average generation tree accuracy (column `acc') over all generated sentences.

The first two rows show results for non-statistical {\sc crisp} and confirm our hypothesis that this problem formulation does not scale up to the automatically induced treebank. The greedy search with {\sc spud} heuristics does not generate a single correct sentence, as the combinatorics of the planning domain becomes too large with our grammar. 
With the standard {\sc ptag} definition our system manages to generate only very few sentences due to data sparseness. However these sentences are very similar to the ones found in the test set. Both generalized models ease the data sparseness problem, but output quality drops. For the unlexicalized model the search problem becomes very hard due to the large planning domain, so that 22\% of the generation problems time out. As expected the smoothed model provides better coverage than {\sc ptag}, while producing higher output quality than the unlexicalized model. 
\footnote{In our experiments we set the weight between level 1 and 2 to 0.9 and did not consider level 3, after empirically testing different settings. We set a probability threshold to include operators into the planning domain to $10^{-4}$}.

\begin{table}
    \caption{\label{results} Results for the realization experiment. Planners: `greedy'= adapted {\sc spud} strategy, `FF' = reimplementation of {\sc ff}, `FF + costs' = cost-sensitive heuristics } 
    \begin{center}
    \begin{tabular}{|l||l||r|r|r|}
    \multicolumn{5}{c}{\sc crisp}\\ 
    \hline
     {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     greedy         & ??? & 0\% & 100\% & 0\%  \\ \hline 
     FF         &  0.66 & 45\% & 42\% & 13\% \\ \hline 
    \multicolumn{5}{c}{\bf planning with {\sc ptag}}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs &  0.90 &  10\% &88\% & 2\% \\ \hline
     FF         &  0.87 & 7\% & 67\% &  26\% \\ \hline
    \multicolumn{5}{c}{\bf planning with the unlexicalized model}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs & 0.74 & 62\% & 16\% & 22\%\\ \hline
     FF & 0.76 & 62\% & 16\% &  22\%\\ \hline
    \multicolumn{5}{c}{\bf planning with linear interpolation}\\\hline
    {\bf Planner }  & {\bf acc}& {\bf success} & {\bf fail} & {\bf timeout} \\ \hline 
     FF + costs & 0.88 & 19\% & 74\% & 7\% \\ \hline
     FF & 0.78 & 19\% & 74\% & 7\% \\ \hline
    \end{tabular}
    \end{center}
\end{table}

### 2010-05-21 trunk/Papers/pcrisp-10/pcrisp.tex
experiment section

Finally, we set the cost of an operator to be its negative log probability. For example for the operator in figure \ref{example-action} $$Cost( \subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2) ) =$$
$$ -\log( P(\subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2)) ). $$ This way the plan with the smallest cost sum corresponds to the {\sc tag} derivation with the highest probability.

>>>>

Finally, we set the cost of an operator to be its negative log probability. For example 
\[
Cost( \subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2) ) =
\]
\[ 
 -\log( P(\subst(t_3,\textit{`cat'},t_{28},\textit{`eats"},n_2)) ). 
\] 
This way the plan with the smallest cost sum corresponds to the {\sc tag} derivation with the highest probability.

### 2010-05-21 trunk/Papers/pcrisp-10/related.tex
Spell checked
experiment section

\section{Related Work}
\label{sec:related}
Statistical methods are extremely popular for surface realization, but have not been used in systems that integrate sentence planning. Most statistical generation approaches follow a generate-and-select strategy, first proposed by \newcite{knight1995} in their {\sc nitrogen} system. Such systems generate a set of candidate sentences using a conventional (non-statistical) grammar and then select the best output sentence by applying a statistical language model. This family includes systems such as {\sc HAL}ogen \cite{langkildeknight1998,langkilde2000} and open{\sc CCG} \cite{whitebaldridge2003}.  The {\sc fergus} system by \newcite{bangalorerambow2000} is a variant of this approach that, like {\sc pcrisp}, employs {\sc ltag}. It first assigns elementary trees to each entity in the input sentence plan using a statistical tree model and then computes the most likely derivation using using only these trees with an $n$-gram model on the output sentence. An alternative to the generate and select approach is to use a probabilistic grammar model, like {\sc ptag}, trained on automatic parses \cite{zhongstent2005}. A related approach uses a model over local decisions of the generation system itself \cite{belz2008}. Both model can either be used to discriminate a set of output candidates, or more directly to choose the next best decision locally. Our approach is similar, because it uses {\sc ptag} to find the most likely output structure. However, the previous work discussed so far addresses surface realization only.
The generate and select approach is even incompatible with the integrated treatment of sentence planning and realization, because it discriminates between possible linearization of the same lexical material. In contrast, {\sc pcrisp} addresses sentence planning as well.

Our treatment of integrated sentence planning and surface realization as planning, is inherited directly from {\sc crisp} \cite{kollerstone2007}.  Planning has long played a role in generation, but has focused on discourse planning instead of specifically addressing sentence generation \cite{hovy1988,appelt1992}. The applicability of these ideas was limited at that time because efficient planning technology was not available. Recently the development of more efficient planning algorithms (e.g. \newcite{hoffmannnebel2001,richterwestphal2008}) spawned a renewed interest in planning for NLG.  
{\sc Crisp} uses such efficient algorithms to efficiently solve the sentence generation problem defined by {\sc spud} \cite{stonedoran1997}. {\sc Spud}, which instead used an incomplete greedy algorithm, is based on an {\sc ltag} whose trees are augmented with semantic and pragmatic constraints. Given a communicative goal, a solution to the {\sc spud} problem realizes this goal and simultaneosly selects refering expressions. The next section explores this idea in more detail.

>>>>

\section{Related Work}
\label{sec:related}
Statistical methods are extremely popular for surface realization, but have not been used in systems that integrate sentence planning. Most statistical generation approaches follow a generate-and-select strategy, first proposed by \newcite{knight1995} in their {\sc nitrogen} system. Such systems generate a set of candidate sentences using a conventional (non-statistical) grammar and then select the best output sentence by applying a statistical language model. This family includes systems such as {\sc HAL}ogen \cite{langkildeknight1998,langkilde2000} and open{\sc CCG} \cite{whitebaldridge2003}.  The {\sc Fergus} system \cite{bangalorerambow2000} is a variant of this approach that, like {\sc pcrisp}, employs {\sc ltag}. It first assigns elementary trees to each entity in the input sentence plan using a statistical tree model and then computes the most likely derivation using using only these trees with an $n$-gram model on the output sentence. An alternative to the generate and select approach is to use a probabilistic grammar model, like {\sc ptag}, trained on automatic parses \cite{zhongstent2005}. A related approach uses a model over local decisions of the generation system itself \cite{belz2008}. Both model can either be used to discriminate a set of output candidates, or more directly to choose the next best decision locally. Our approach is similar, because it uses {\sc ptag} to find the most likely output structure. However, the previous work discussed so far addresses surface realization only.
The generate and select approach is even incompatible with the integrated treatment of sentence planning and realization, because it discriminates between possible linearization of the same lexical material. In contrast, {\sc pcrisp} addresses sentence planning as well.

Our treatment of integrated sentence planning and surface realization as planning, is inherited directly from {\sc crisp} \cite{kollerstone2007}.  Planning has long played a role in generation, but has focused on discourse planning instead of specifically addressing sentence generation \cite{hovy1988,appelt1992}. The applicability of these ideas was limited at that time because efficient planning technology was not available. Recently the development of more efficient planning algorithms (e.g. \newcite{hoffmannnebel2001,richterwestphal2008}) spawned a renewed interest in planning for NLG.  
{\sc Crisp} uses such efficient algorithms to efficiently solve the sentence generation problem defined by {\sc spud} \cite{stonedoran1997}. {\sc Spud}, which instead used an incomplete greedy algorithm, is based on an {\sc ltag} whose trees are augmented with semantic and pragmatic constraints. Given a communicative goal, a solution to the {\sc spud} problem realizes this goal and simultaneously selects referring expressions. The next section explores this idea in more detail.
